Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-07-28 resubmission/04-analisis.qmd
Update varios ajustes

```
# Results

@@

According to Model 1, which includes the survey waves to capture intertemporal variations in the dependent variable, there is a decrease in 2017 ($\beta$ = -0.167, $p$ < .001) relative to 2016, and similarly in 2018 ($\beta$ = -0.025, $p$ > .05) and 2019 ($\beta$ = -0.046, $p$ > .05), although the latter effects are not statistically significant. In contrast, in the more recent waves of 2022 and 2023, there is a statistically significant increase in market justice preferences ($\beta$ = 0.285, $p$ < .001; $\beta$ = 0.284, $p$ < .001), suggesting a non-linear effect. To model this trajectory over time, Model 2 incorporates time (survey waves) as a continuous variable, along with its quadratic term, representing the non-linear association initially observed in Model 1. While the linear term (survey wave) shows a negative association, reflecting an overall decline in market preferences over time, the positive quadratic term indicates a reversal of this pattern in the final measurement points.

Models 3 and 4 incorporate the within-group effects (WE) of the primary independent variables, capturing how individual changes in these variables over time shape the dependent variable. The results in Model 3 suggest that the within effect of perceived economic inequality is negative and statistically significant ($p$ < .001). Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.039 point decrease in market justice preferences. Model 4 shows that meritocratic perceptions operate in the opposite direction. An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.070, $p$ < .001), and a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with higher market-justice preferences ($\beta$ = 0.027, $p$ < .05). Taken together, these results suggest that individuals who increasingly perceive meritocracy—whether through effort or talent—tend to hold stronger market justice preferences.

When examining the between-group effects (BE) in Model 5 and 6, which capture differences between individuals in the average of the main variables, a similar pattern emerges. Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.042, $p$ < .05). In Model 6, the meritocratic perception that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.186, $p$ < .001), whereas the perception that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.040). Notably, once the meritocratic variables are included, the negative coefficient for perceived economic inequality remains but is no longer statistically significant.

Model 7 adds the control variables. With the exception of the between-effect for perceived economic inequality—which becomes nonsignificant—the within- and between-effects of the principal predictors retain both their direction and statistical significance, confirming the robustness of the associations (see Supplementary Material for effects of control variables).

@@

@tbl-interactions1 examines whether perceived economic inequality moderates the effect of meritocratic perceptions on market justice preferences. Contrary to our hypothesis, the interaction terms in the within-person specification (Models 8 and 9) do not reach statistical significance. However, in the between-person specification (Models 10 and 11), the interaction terms are statistically significant, suggesting that perceptions of economic inequality meaningfully alter the effect of meritocracy on support for market-based allocation of social services. Specifically, these positive effects indicate that, as perceptions of inequality increase, the positive association between meritocratic perceptions and market justice preferences becomes stronger. As shown in @fig-interact, Panel A, the effect of meritocratic perceptions about effort on market justice preferences strengthens as perceived inequality increases, particularly among individuals who perceive the highest levels of inequality, whereas this effect weakens among those who perceive low levels of inequality. Panel B reveals a similar pattern for meritocratic perceptions about talent, although the effect is substantially weaker at low levels of perceived inequality. Substantively, these findings suggest that the legitimizing power of meritocratic perceptions is amplified in contexts perceived as highly unequal, reinforcing individualistic understandings of justice and strengthening support for market-based access to essential social services.

@@

With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time only has significant effects on the between effects of the meritocratic variables. These interaction effects are negative and mean that in the last waves, the impact of meritocratic variables on market justice preferences is attenuated. Specifically, as time advances, the positive within-person effect of perceiving that effort is rewarded (Model 16) or that intelligence is rewarded (Model 17) progressively attenuates. Put differently, the uplift in market justice preferences associated with a strong meritocratic perception becomes smaller in each successive wave.

>>>>

# Results

@@

According to Model 1, which includes the survey waves to capture intertemporal variations in the dependent variable, there is a decrease in 2017 ($\beta$ = -0.183, $p$ < .001) relative to 2016, and similarly in 2018 ($\beta$ = -0.009, $p$ > .05) and 2019 ($\beta$ = -0.009, $p$ > .05), although the latter effects are not statistically significant. In contrast, in the more recent waves of 2022 and 2023, there is a statistically significant increase in market justice preferences ($\beta$ = 0.300, $p$ < .001; $\beta$ = 0.320, $p$ < .001), suggesting a non-linear effect. To model this trajectory over time, Model 2 incorporates time (survey waves) as a continuous variable, along with its quadratic term, representing the non-linear association initially observed in Model 1. While the linear term (survey wave) shows a negative association, reflecting an overall decline in market preferences over time, the positive quadratic term indicates a reversal of this pattern in the final measurement points.

Models 3 and 4 incorporate the within-group effects (WE) of the primary independent variables, capturing how individual changes in these variables over time shape the dependent variable. The results in Model 3 suggest that the within effect of perceived economic inequality is negative and statistically significant ($p$ < .001). Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.027 point decrease in market justice preferences. Model 4 shows that meritocratic perceptions operate in distinct directions. An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.070, $p$ < .001), while a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with lower market-justice preferences ($\beta$ = -0.027, $p$ < .05). Taken together, these results suggest that people who increasingly perceive meritocracy based on effort tend to have stronger preferences for market justice, and that the opposite is true for those who increasingly perceive meritocracy based on talent.

When examining the between-group effects (BE) in Model 5 and 6, which capture differences between individuals in the average of the main variables, a similar pattern emerges. Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.002, $p$ > .05). However, this effects is no longer statistically significant. In Model 6, the meritocratic perception that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.206, $p$ < .001), whereas the perception that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.036).

Model 7 adds the control variables. The within- and between-effects of the principal predictors retain both their direction and statistical significance, confirming the robustness of the associations (see Supplementary Material for effects of control variables).

@@

@tbl-interactions1 examines whether perceived economic inequality moderates the effect of meritocratic perceptions on market justice preferences. Contrary to our expectations, the interaction terms in the within-person specification of Model 9 indicate that the negative effect of talent-based meritocratic perceptions becomes stronger as perceived inequality increases ($\beta$ = -0.032, $p$ < .05). In contrast, in the between-person specification (Model 11), the interaction term is positive ($\beta$ = 0.099, $p$ < .01), suggesting that perceptions of economic inequality significantly shape the influence of meritocratic perceptions on support for market-based allocation of social services. Specifically, these positive effects indicate that, as perceptions of inequality increase, the association between talent-based meritocratic perceptions and market justice preferences shifts in a positive direction.

The effects of effort-based meritocratic perceptions are not statistically significant in either the within- or between-person specifications (Models 8 and 10). However, the direction of the coefficients is negative in the former and positive in the latter. As shown in @fig-interact, Panel A, the between-person effect of effort-based meritocratic perceptions on market justice preferences strengthens as perceived inequality increases across individuals, whereas this effect weakens among those who perceive lower levels of inequality. Panel B reveals a similar pattern for the between-person effect of talent-based meritocratic perceptions, although the effect is substantially weaker at low levels of perceived inequality.

Substantively, these findings suggest that the legitimizing function of meritocratic perceptions is amplified in contexts perceived as highly unequal. In such contexts, meritocratic narratives may reinforce individualistic understandings of distributive justice and intensify support for market-based access to essential social services.

@@

With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time has no statistically significant effects on either the within- or between-person effects of perceived inequality and meritocratic perceptions. The within-person interaction effects are positive, suggesting that in later waves, the influence of perceived inequality and meritocratic beliefs on market justice preferences tends to increase. Conversely, in the between-person specification, the interaction effects are negative, indicating that over time, the impact of perceived inequality and meritocratic beliefs on support for market-based justice tends to diminish. However, these trends are not statistically significant, and thus no definitive conclusions can be drawn.
